\textbf{Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota}

### 2017-06-12 submission/manuscript.tex
first draft for editing done

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\begin{document}

\maketitle

@@

\newpage
\linenumbers

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide.
Tumors of the proximal (right) and distal (left) colon are
morpologically and genetically distinct. Previous work from our group
found that microbial dysbiosis is associated with the development of
colorectal cancer tumors in studies of both mice and humans. Analysis of
the fecal microbiota from healthy and CRC patients further revealed
different microbial signatures associated with disease. In this study,
we extended our observations of the fecal microbiome to analysis of the
proximal and distal human colon. We used a two-colonoscope approach on
subjects that had not undergone standard bowel preparation procedure.
This technique allowed us to characterize the native proximal and distal
luminal and mucosal microbiome without prior chemical disruption. 16S
rRNA gene sequencing was performed on proximal and distal mucosal
biopsies, luminal and exit stool for 20 healthy individuals. Diversity
analysis of each location revealed that each site contained a diverse
community, and that a patient's samples were more similar to each other
than to that of other individuals. Since we could not differentiate
sites along the colon based on community structure or community
membership alone, we employed the machine-learning algorithm Random
Forest to identify key species that distinguish biogeographical sites.
Random Forest classification models were built using taxa abundance and
sample location revealed distinct populations that were found in each
location. Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas
and Actinomyces were most likely to be found in mucosal samples versus
luminal samples (AUC = 0.925). The classification model performed well
(AUC = 0.912) when classifying mucosal samples into proximal or distal
sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The left mucosa was found to have high
populations of Finegoldia, Murdochiella and Porphyromonas. Proximal and
distal luminal samples were comprised of many of the same taxa, likely
reflecting the fact that stool moves along the colon from the proximal
to distal end. Finally, comparison of all samples to fecal samples taken
at exit uncovered that the feces were most similar to samples taken from
the left lumen, again reflecting the anatomical structure of the colon.
Taken together, our results have identified distinct bacterial
populations distinct of the proximal and distal colon. Further
investigation of these bacteria may elucidate if and how these groups
contribute to differential oncogenesis processes on the respective sides
of the colon.

@@

\begin{itemize}
\item
  colon cancer
\item
  right and left tumors are morphologically and genetically distinct
  (why Q is important)
\item
  the microbiome of healthy people and patients with CRC is
  distinguishable at the OTU level
\item
  specific keystone species are associated with CRC tumors
\item
  Central Question: do the populations of bacteria in the left and right
  colon contribute to the development of specific subtypes of CRC?
\item
  nearly all studies of the gut microbiome have been from shed stool or
  colonoscopy with bowel prep
\item
  bowel prep changes the mucosal/fecal microbiome and what can be
  analyzed from it.
\item
  how does the microbiome vary spatially along the colon? we know that
  the environment varies dramatically but we dont know how those
  environmental factors specifically affect bacterial populations
\item
  What data is needed to answer this question? first, need to
  characterize healthy species that are specific to each location. then
  we can ask, if/how disruption affects tumorigenesis? so what is needed
  is paired samples from healthy people that have not undergone bowel
  prep, from the right/left and lumen/mucosa
\item
  what methods are used to get the data? unprepped colonoscopy
\item
  What analysis must be applied? - diversity analysis and machine
  learning to pull out small variations/differences among the noisy data
  set.
\item
  hypothesis: there are distinct bacteria on the left and right side
  that can be distinguished by 16S sequencing of stool
\end{itemize}

@@

\begin{itemize}
\item
  what data were obtained? 16S sequencing results for samples taken from
  healthy human colons that had not undergone bowel prep Figure 1:
  sample collection schematic
\item
  what were the results of the analyses? the results of the analyses are
  as follows:
\end{itemize}

\paragraph{Descriptive analysis}\label{descriptive-analysis}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  there are similar levels (within error margins) of each phyla for each
  location, but there is noise due to interpatient variability. Overall,
  there is also a similar level of diversity in each location. However,
  each patient has a different pattern of diversity over the locations,
  some are high in R, some are high in L, etc. Figure 2:
\item
  2A: relative abundance by phyla
\item
  2B: simpson diversity
\item
  2C: NMDS/ordination of some kind
\end{itemize}

@@

-how did the analyses answer the central question?

-what does this answer tell us about the broader field?

@@

@@

Study participants/IRB

Sample collection

Sample processing/DNA extraction

16S rRNA sequencing

Statistical analysis/Random Forest

\subsubsection{Figure Legends}\label{figure-legends}

>>>>

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Machine learning identifies unique taxa differentiating proximal and distal human colonic microbiota},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{document}
\maketitle

@@

\newpage

\linenumbers

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide.
Tumors of the proximal (right) and distal (left) colon are
morphologically and genetically distinct. Previous work from our group
found that microbial dysbiosis is associated with the development of CRC
tumors in studies of both mice and humans. Analysis of the fecal
microbiota from healthy and CRC patients further revealed different
microbial signatures associated with disease. In this study, we extended
our observations of the fecal microbiome to analysis of the proximal and
distal human colon. We used a two-colonoscope approach on subjects that
had not undergone standard bowel preparation procedure. This technique
allowed us to characterize the native proximal and distal luminal and
mucosal microbiome without prior chemical disruption. 16S rRNA gene
sequencing was performed on proximal and distal mucosal and luminal
biopsies and exit stool for 20 healthy individuals. Diversity analysis
revealed that each site contained a diverse community, and that a
patient's samples were more similar to each other than to that of other
individuals. Since we could not differentiate sites along the colon
based on community structure or community membership alone, we employed
the machine-learning algorithm Random Forest to identify key species
that distinguish biogeographical sites. Random Forest classification
models were built using taxa abundance and sample location and revealed
distinct populations that were found in each location.
\emph{Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas} and
\emph{Actinomyces} were most likely to be found in mucosal samples
versus luminal samples (AUC = 0.925). The classification model performed
well (AUC = 0.912) when classifying mucosal samples into proximal or
distal sides, but separating luminal samples from each side proved more
challenging (AUC = 0.755). The left mucosa was found to have high
populations of \emph{Finegoldia, Murdochiella} and \emph{Porphyromonas}.
Proximal and distal luminal samples were comprised of many of the same
taxa, likely reflecting the fact that stool moves along the colon from
the proximal to distal end. Finally, comparison of all samples to fecal
samples taken at exit uncovered that the feces were most similar to
samples taken from the left lumen, again reflecting the anatomical
structure of the colon. Taken together, our results have identified
distinct bacterial populations of the proximal and distal colon. Further
investigation of these bacteria may elucidate if and how these groups
contribute to differential oncogenesis processes on the respective sides
of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related
deaths in the United States. CRC tumors vary in structure, size,
morbidity and symptomology depending upon their geographical location in
the colon. CRC has recently been described to fall into four molecular
subtypes ((1)). Of these subtypes, tumors that arise on the left side of
the colon are infiltrating lesions that present with painful symptoms.
In contrast, right-sided colon tumors are bulky and project into the
lumen, often causing no symptoms until they have developed into
carcinomas ((2)). Because of the absence of symptoms, right-sided
tumors, which comprise up to 47\% of CRCs, have a significantly higher
mortality rate ((2)). The left and right sides of the colon differ in
the amount of inflammation present and the genomic instability of
precancerous cells, respectively, as well as oxygen, pH and the presence
of antimicrobial peptides ((3), (4), (5)). Microenvironments differ not
only longitutinally along the colon, but latitudinally from the
epithelium to mucosa to intestinal lumen, offering several sites for
different microbial communities to flourish. Given this varied
physiology of the proximal-distal axis of the colon, symbiotic microbes
and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of
CRC can be attributed to specific molecular events as a result of
interactions between the gut microbiota and human host ((3)). Our group
and others have found that the stool microbiome of patients with CRC is
distinct from that of healthy people ((6)). Further studies manipulating
the gut microbiome using antibiotics or other chemoagents in mice has
shown that dysbiosis preceded and accelerated the development of CRC
tumors ((7)). Comparison of the specific bacteria present on CRC tumors
with those found on nearby healthy tissue has also identified specific
bacterial species that are tumor-associated based on FISH staining and
16S rRNA sequencing ((8)). These species include the oral pathogens
\emph{Fusobacterium nucleatum} and \emph{Porphyromonas asacharolytica}.
Interestingly, the periodontal pathogens are also highly predictive of
if a patient has CRC tumors or not in our classification studies ((9)).
However, these studies have been performed comparing shed stool or in
mice, which does not allow for analysis of paried samples from the
proximal and distal sides of the colon. Similarly, comparisons of on- or
off-tumor bacteria rarely have matched tissue from the other side of the
colon from the same patient, limiting what conclusions can be drawn
about the colonic microbiome overall, let alone at that specific site.
Disruption of the healthy community could promote the initiation or
proliferation of the distinct left and right CRC tumors. However, the
contribution of the gut microbiota to these subtypes is largely
undefined.

Furthermore, profiles of the microbial biogeography of the gut have been
limited by sample collection methods. The majority of human gut
microbiome studies have been performed on whole shed stool or on samples
collected during colonoscopy procedures. While the latter method allows
investigators to acquire samples from inside the human colon, typically
this procedure is preceded by the use of bowel preparation methods such
as the consumption of laxatives to cleanse the bowel. Bowel preparation
is essential for detecting cancerous or precancerous lesions in the
colon, but complicates microbiome profiling as the chemicals strip the
bowel of contents and disrupt the mucosal layer ((10), (11)). As such,
what little information we do have about the biogeographical
distribution of the microbes in the proximal and distal colon is
confounded by the bowel preparation procedure.

In this study we sought to combine these approaches to characterize the
microbiome in the lumen and mucosa of the proximal and distal healthy
human colon. Our design uses unprepared colonoscopy techniques to sample
each location of the gut without prior disruption of the native flora in
20 healthy volunteers. To get around the noise created by a diverse set
of human microbiomes, we used a machine-learning classification
algorithm trained on curated 16S rRNA sequencing reads to identify
microbes specific to each location. We found that our classification
models were able to separate mucosal and lumenal samples as well as
differentiate between sides of the colon based on populations of
specific microbes. By identifying the specific microbes we are poised to
ask if and how the presence or disruption of the microbes at each site
contribute to the development of the specific tumor subtypes of CRC in
the proximal and distal human colon.

@@

\paragraph{Microbial membership and diversity of the proximal and distal
colon}\label{microbial-membership-and-diversity-of-the-proximal-and-distal-colon}

Proximal and distal lumenal and mucosal samples were collected from 20
healthy humans that had not undergone bowel preparation(Figure 1).
Participants also collected stool at home one week prior to the
procedure. To characterize the bacterial communities present at these
sites, 16S rRNA gene sequencing was performed on extracted DNA from each
sample. The relative abundance of each phyla for each site is shown in
Figure 2A. This data combines the samples from all participants in the
study. Each site is primarily dominated by \emph{Firmicutes} and
\emph{Bacteriodetes}, consistent with known variability in human
microbiome research. Likewise, samples had varying levels of diversity
at each site, irrespective of the individual (Figure 2B). That is, while
some individuals had a more diverse right mucosa, some had a more
diverse left mucosa, and so on. Therefore we could not identify a clear
pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or
mucosa), we calculated theta YC distances from OTU abundances and
compared distances for all individuals. Again, across all patient
samples we saw a range of theta YC distances when comparing sample
locations (Figure 3A) and again those ranges did not follow a clear
pattern on an individual basis. However, when comparing median distances
between the right lumen and mucosa, the left versus right lumen, the
left versus right mucosa and the left lumen and mucosa, we found that
the right lumen and mucosa were most similar to each other than the
other samples (P \textless{} 0.005, Wilcoxon, BH adjustment). Next, we
calculated thetaYC distances to examine how each sample compared to the
home-collected exit stool. Amidst variability between patients, we did
identify significantly smaller thetaYc distance between the left lumenal
sample and the exit stool (Figure 3B, P \textless{} 0.05, Wilcoxon, BH
adjustment). Further, there was an even larger difference in the
comparisons of the left mucosa to the exit stool, indicating that the
mucosa is quite different from the stool as compared to lumen (P
\textless{} 0.0005, Wilcoxon, BH adjustment). To determine what factors
may be driving the differences seen among the samples, we compared
thetaYC distances between samples from all subjects (interpersonal)
versus samples from within one subject (intrapersonal). We found that
samples from one individual were overall much more similar to each other
than to other study subjects (Figure 3C), consistent with previous human
microbiome studies that have sampled multiple sites of the human colon
(\textbf{???}). Thus interpersonal variation between subjects drives the
differences between samples more than sample site or location. Overall,
the results of the similarity analysis suggest that the contents of the
left lumen are most representative of stool at exit, and the microbes
remaining on the mucosa are much different.

\paragraph{Random Forest classification models identify important OTUs
on each
side}\label{random-forest-classification-models-identify-important-otus-on-each-side}

To identify OTUs that were distinct at each biogeographical site, we
constructed several Random Forest machine-learning models built on OTU
abundances. The first model we built to classify the lumen versus
mucosal samples for the right and left sides, independently (Figure 4A).
The constructed model used ((Xopt)) features for the right and ((Xopt))
for the left. The models performed well when classifying these samples
(AUC = 0.797 and AUC = 0.980, respectively). The OTUs that were most
predictive of each site are identified by their greatest mean decrease
in accuracy when removed from the model. For distinguishing the right
lumen and mucosa, OTUs from the \emph{Bacteriodes}, \emph{Actinomyces},
\emph{Psuedomonas} and two OTUs from the \emph{Enterobacteraceae} genera
were differentially abundant (Figure 4B). The model classifying the left
lumen and mucosa identified OTUs from \emph{Turicibacter},
\emph{Finegoldia}, \emph{Peptoniphilus} and two OTUs from the
\emph{Anaerococcus} genera that could distinguish lumen from mucosal
samples (Figure 4C). These results indicate that there are fine
differences between the different sites of the colon, and that these can
be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome.
The model performed best when distinguishing the right versus left
mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen
(AUC = 0.755). The model was able to explain ((X\%)) of the variance.
OTUs that were differentially abundant between the left and right mucosa
included members of the \emph{Porphyromonas}, \emph{Murdochiella},
\emph{Finegoldia}, \emph{Anaerococcus} and \emph{Peptoniphilus} genera
(Figure 5B). The model was less-effective at classifying the right and
left lumen (AUC = 0.755). However it did identify differentially
abundant OTUs such as three members of the \emph{Bacteroides} genus, a
\emph{Clostridium IV} OTU and an \emph{Oscillibacter} OTU (Figure 5C).
This analysis found that some of the same OTUs that are distinct between
the mucosa and lumen also drive sidedness- such as \emph{Anaerococcus}
and \emph{Finegoldia}.

\paragraph{Bacterial OTUs associated with cancer are found in healthy
individuals}\label{bacterial-otus-associated-with-cancer-are-found-in-healthy-individuals}

Given that specific bacterial species have been associated with
colorectal cancer we probed our sample set for these OTUs. Among our 100
samples, the most frequent sequence associated wtih the
\emph{Fusobacterium} genus was OTU179, which BLASTs to
\emph{Fusobacterium nucleatum subsp animalis}. It was the only sequence
to align with a \emph{F. nucleatum} strain- the only species of
\emph{Fusobacterium} known to have oncogenic properties and be found on
the surfaces of colorectal cancer tumors. ((12)). The
\emph{Fusobacterium} positive samples were located on the right and left
mucosa and ranged up to 1\% of the sample (Figure 6A). OTU00152 is
\emph{Porphyromonas} and the most frequent sequence in that OTU aligns
to \emph{Porphyromonas asacharolytica}, another bacterium commonly
detected and isolated from colorectal tumors. OTU152 was only detected
on the left mucosa, and in fact was one of the OTUs the classification
model identified as separating left and right sides (Figure 6B). Samples
that were positive for \emph{P. asacharolytica} ranged in abundances
from 0.01\% - 16\%. Thus, cancer-associated OTUs could be found in our
sample set of 20 healthy individuals.

@@

Here we identified bacterial taxa that are specific to the lumen and
mucosa of the right and left human colon from samples collected during
unprepared colonoscopy. We found that all locations contained a range of
phyla and a range of diversity, but that there was a wide variability
between subjects. Pairwise comparisons of each of the sites revealed
that the right mucosa and lumen were most similar to each other.
Further, comparison of colonoscopy-collected samples with samples
collected from stool at home showed that the left lumen is most similar
to stool at exit. Random Forest algorithms built on OTU abundances from
each sample identified microbes that are particular to each location of
the colon. Finally, we were able to detect some bacterial OTUs
associated with colorectal cancer in our healthy patient cohort. Using
unprepped colonoscopy and machine learning, we have identified bacterial
phyla specific to the healthy proximal and distal human colon.

When examining the relative abundance of the different phyla at each
site, there is a wide amount of variation for each phyla with
communities primarily dominated by the Bacteriodes and Firmicutes. This
likely reflects not only the variability between human subjects, casued
by differences in age, gender, diet, but also reports of microbial
``patchiness'' in the gut microbiome. Several previous studies have
noted that the bacteria recoverable from the same mucosal sample
location can be vastly different when the samples are taken just 1 cm
away from each other ((13)). Similar patchiness is also observed in
lumenal contents and fecal samples themselves; there is observed
separation of different interacting microbes along the length of a stool
sample, for instance ((14)). That said, across our samples the mucosal
samples harbor more Proteobacteria, consistent with previous studies
comparing mucosal swabs to lumenal content in humans ((4)). Hence, the
conclusions we can draw from phyla analysis are likely confounded by
differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and
between each anatomical site. As this measurement takes both number of
species and abundance of species into account, it seems that there is a
varied level of each component along the human colon. There was no clear
pattern of diversity when comparing patients- some patients had a higher
level of diversity in the proximal than the distal colon, some patients
had the opposite. An early study of microbial diversity and membership
of three human subjects found a varied level of diversity in each
subject, but that subjects were overall more similar to themselves than
to each other (\textbf{???}). We used the thetaYC dissimilarity metric
to compare the communities present at each anatomical location to each
other. We found that while there was a wide range of similarity of each
site within a person, the study participants were ultimately more
similar to themselves than all of the other participants. This finding
is consistent with numerous studies examining the biogeography of the
colon in healthy humans (\textbf{???}, (15), (16)).

To get around the noisiness created by a diverse set of samples, we
built a Random Forest model to identify microbes specific to each side.
For each comparison we identified top X OTUs that were strongly
predictive of one site or another. Generally, OTUs identified in each
location were consistent with known physiological gradients along the
gut axis ((5)). For instance, the right mucosa harbored
mucosa-associated facultative anaerobes such as \emph{Actinomyces} and
\emph{Enterobacteraceae} and the aerobic \emph{Psuedomonas} consistent
with the highest oxygen regions of the colon. The left mucosa was far
more likely to host strictly anaerobic species such as
\emph{Porphyromonas}, \emph{Anaerococcus}, \emph{Finegoldia} and
\emph{Peptoniphilus}. The model was less effective at classifying the
right and left lumenal contents, probably because the samples are
arguably composed of the same material.

We detected \emph{F. nucleatum} and \emph{P. asacharolytica} in 7 and 5
of our subjects, respectively. These bacteria have been shown to be
predictive of colorectal cancer in humans ((9)) and have oncogenic
properties in cell culture and in mice ((17)). Interestingly, while
\emph{F. nucleatum} was found on both sides of the colon, \emph{P.
asacharolytica} was only detected in the left mucosa. Not much is known
about the distribution of \emph{P. asacharolytica} but given its
documented anaerobic characteristics and asacharolytic metabolism, it
might not be surprising that it resides in the less-oygen-rich and
proteinaceous left mucosa ((4)). In studies examining bacteria on
colorectal cancer tumors, \emph{F. nucleatum} is more commonly detected
on right-sided tumors, and distribution of \emph{F. nucleatum} decreases
along the colon to rectum ((18)). Of the (8) (40\%) individuals positive
for \emph{F. nucleatum} in this study, the bacteria was spread across
the right mucosa, left lumen and left mucosa. That said, in a previous
study of 1,102 colorectal cancer cases, only 138 (11\%) of them
contained detectable \emph{F. nucleatum} at all ((18)). In a separate
study of patients with inflammatory bowel disease (IBD), 50\% of 22 IBD
patients and 17.6\% of 34 people without disease contained \emph{F.
nucleatum} in mucosal biopsies ((19)). These studies have the added
caveat of being sampled from prepared or cleaned colons. Thus, the
presence of \emph{F. nucleatum} in a healthy individual is not
necessarily linked to the development of future colorectal cancers.
Because of the spatial distribution of the \emph{F. nucleatum} in our
sample set, we cannot develop a model for the role of \emph{F.
nucleatum} in the healthy colon. Indeed, rather than the \emph{F.
nucleatum} initiating tumorigenesis, these bacteria may instead arrive
``late'' to the tumor microenvironment and feed on tumors that have
alread begun to develop ((20)). Data examining bacterial biofilms on CRC
tumors suggests that \emph{Fusobacteria} species are more commonly found
both on proximal tumors and in biofilms, indicating that it is not only
the presence of the bacteria but the organization of the tumor community
that contributes to \emph{Fusobacterium's} role in tumorigenesis.
Finally, \emph{Fusobacteria} and \emph{Porphyromonas} species have been
known to not only co-occur on CRC tumors but also to synergistically
promote tumorigenesis in an oral cancer model ((21), (20)). Thus,
further analysis of the distribution and role of these pathogens may
elucidate a mechanism for development of CRC tumors in the proximal or
distal colon.

Specific comparisons of our findings to previously published gut
biogeography studies are additionally confounded by the use of bowel
preparation methods in most other studies. A rare report of a
matched-colonoscopy study that sampled 18 patient's colonic mucosa and
lumenal contents prior to and after bowel cleansing ((22)). This group
found that mucosa and lumenal samples were distinguishable prior to
bowel cleansing, but that bowel preparation resulted in an increase in
shared OTUs between each site ((22)). Bowel cleansing not only made the
samples harder to distinguish, it resulted in decreases in diversity
across sites. Further, the differences were not great enough to overcome
interpersonal differences between subjects. So, bowel preparation
clearly induces bias into the microbes recovered from sampling the lumen
or mucosa of a prepared bowel. Thus our findings of specific bacteria at
each site of the colon are strengthened by the lack of bowel
preparation.

Microbiome-based diagnostics are increasingly being explored as
non-invasive tools to survey for the development of colon cancer. Random
Forest models have been used by our group and others to increase the
detection sensitivity of CRC tumors. Indeed, our group found that a
classification model that used microbiome data in combination with Fecal
Immunohistochemical Test (FIT) results could correctly identify both
carcinoma and adenoma lesions from communities of stool at exit and it
performed much better than FIT alone. Further work from our lab has
shown that microbiome profiling of the FIT cartridge contents
sufficiently represented the stool community ((9)). One caveat of the
FIT study was that there was not sufficient information to test if a
classification model could differentiate between proximal and distal CRC
tumors based on exit stool sample alone, but we would hypothesize that
would not be effective. Given that our results showed that the stool
most accurately reflects the community of the left lumen, we likely
cannot use Random Forest of stool samples to diagnose any changes in the
proximal or mucosal communities.

By revealing specific differences in microbial populations at each
location in the gut via sampling an unprepared bowel, we can begin to
form hypothesies about how specific host-microbe interactions can affect
oncogenesis of proximal and distal CRC tumors. To this point, 16S
community profiling studies do not provide enough information to probe
these questions. Our sample set of matched proximal, distal, lumenal and
mucosal samples from colons that have not undergone bowel preparation
presents a unique opportunity to explore further questions about the
microbiome along the gut axis. Specifically, examining metagenomic,
metabolomic and host interactions at each site will be useful in further
characterizing the host-microbe interactions in the development of
proximal and distal colorectal cancer.

@@

We thank all the individuals who volunteered for the study. This work
was supported by the Rose and Lawrence C. Page Foundation (DKT). We
would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
for their roles in study coordination, subject recruitment, sample
collection and sample processing.

@@

\paragraph{Human subjects}\label{human-subjects}

The procedures in this study and consent were approved by the
Institutional Review Board at the University of Michigan Health System
with protocol number XXXX. Subjects were recruited using the online
recruitment platform and were pre-screened prior to enrollment in the
study. Exclusion criteria included: use of asprin or NSAIDs within 7
days, use of antibiotics within 3 months, current use of anticoagulants,
known allergies to Fentanyl or Benadryl, prior history of colon disease,
diabetes, abdominal surgery, respiratory, liver, kidney or brain
impairments, undergoing current chemotherapy or radiation treatment and
subjects that were pregnant or trying to conceive. 20 subjects that met
the criteria were selected and provided signed informed consent prior to
the procedure. There were 13 female and 7 male subjects ranging in age
from 25 to 64.

\paragraph{Sample collection}\label{sample-collection}

At a baseline visit, subjects were consented and given a home collection
stool kit (Source of kit supplies). At least one week prior to the
scheduled colonoscopy, subjects were to collect whole stool at home and
two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.)
and ship the samples to the University. Notably, subjects did not
undergo any bowel preparation method prior to sampling. On procedure
day, subjects reported to the Michigan Clinical Research Unit at the
University of Michigan Health System. Patients were consciously sedated
using Fentanyl, Versed and/or Benadryl as appropriate. A flexible
sigmoidoscope was first inserted about 25cm into the colon and endoscopy
brush used to collect lumenal/stool contents. Two lumenal samples were
collected and the contents immediately deposited into RNAlater (source)
and flash-frozen in liquid nitrogen. The brushes were withdrawn and
biopsy forceps were used to collect mucosal biopsies on sections of the
colon that were pink and free of stool matter. Three mucosal biopsies
were collected and flash-frozen in RNAlater. These samples comprised the
distal or left colon samples. The sigmoidoscope was then withdrawn and a
pediatric colonoscope was inserted to reach the ascending colon. Samples
were then collected as in the distal colon and the colonoscope
withdrawn. All samples were stored at -80 C until study completion.

\paragraph{Sample processing, sequencing and
analysis}\label{sample-processing-sequencing-and-analysis}

DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation
Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP
solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were
added to the bead beating step to enhance DNA recovery from mucosal
samples. The resulting DNA was used as template for amplification of the
V4 region of the 16S rRNA gene and fragments were sequenced on an
Illumina MiSeq as previously described ((23)). Sequences were curated
using the mothur software as described previously ((24)). The sequences
were assigned taxonomic classification using a naive Bayesian classifier
trained using a 16S rRNA gene training set from the Ribosomal Database
Project (RDP) ((25)) and clustered into operational taxonomic units
(OTUs) based on a 97\% similarity cutoff. Sequencing and analysis of a
mock community revealed the error rate to be X\%. Samples were rarefied
to 4231 sequences per sample in order to reduce sampling bias.

Diversity analysis was performed using the Simpson diversity calculator
and theta YC calculator metrics in mothur ((24)). ThetaYC distances were
calculated to determine the dissimilarity between two samples. Random
Forest classification models were built using the randomForest R package
and resultant models were used to identify the OTUs that were most
important for classifying each location ((26)). To get species-level
information about sequences of interest, sequences were aligned using
blastn and the species name was only used if the identity score was
\textgreater{}= 99\%.

\paragraph{Statistical analysis}\label{statistical-analysis}

Differences in community membership at the phyla level were tested using
the analysis of molecular variance (AMOVA) metric in mothur. Differences
in thetaYC distances by location were tested using the Wilcoxon rank-sum
test adjusted for multiple comparisons using the Benjamini-Hochberg
procedure.

\subsubsection{Figure legends}\label{figure-legends}

\paragraph{Figure 1}\label{figure-1}

Sampling strategy. A flexible sigmoidoscope was used to sample the
distal colonic luminal contents and mucosa. The scope was inserted
\textasciitilde{} 25cm into the subject and endoscopy brushes were used
to sample the luminal contents (green star). A separate set of biopsy
forceps was used to sample the distal mucosa (blue star). The
sigmoidoscope was removed. A pediatric colonoscope was inserted and used
to access the proximal colon. Biopsies were taken of the proximal
luminal contents and mucosa as described. One week prior to the
procedure stool was collected at home and sent into the laboratory.
Representative images from one individual are shown.

\paragraph{Figure 2}\label{figure-2}

Microbial membership and diversity of the proximal and distal human
colon. A) Relative abundance of the top five bacterial phyla in each
sampling site. Each box represents the median and confidence intervals.
B) Simpson diversity of the microbial communities at each location. The
lines represent the median values.

\paragraph{Figure 3}\label{figure-3}

Similarity of microbial community structure between sites of the gut.
ThetaYC distances are shown for interpersonal similarities between two
sites -- each point represents one individual. In (A), comparisons of
the right and left mucosal and lumen are shown. In (B), comparisons of
each site to the exit stool are shown.

\paragraph{Figure 4}\label{figure-4}

Random Forest classifies the mucosa and lumen of each side of the colon.
A) Receiver Operator Characteristic curves are shown for the 10-fold
cross validation of the Random Forest model classifying lumen and
mucosal samples for the left and right sides of the colon. (B) Top five
OTUs that are most important for the classification model for the left
mucosa and lumen (B) and the right mucosa and lumen (C).

\paragraph{Figure 5}\label{figure-5}

Random Forest classifies the left and right sides of the colon. A)
Receiver Operator Characteristic curves are shown for the 10-fold cross
validation of the Random Forest model classifying left lumen versus
right lumen (orange) and left mucosa vs right mucosa (green). (B) Top
five OTUs that are most important for the classification model for the
left and right mucosa (B) and the left and right lumen (C).

\paragraph{Figure 6}\label{figure-6}

Location and abundance of cancer-associated OTUs. Relative abundance was
calculated and plotted by sample site for each OTU of interest: (A)
\emph{Fusobacterium nucleatum} and (B) \emph{Porphyromonas
asacharolytica}

\subsubsection*{References}\label{references}
\addcontentsline{toc}{subsubsection}{References}
